---
title: "Human and economic costs of severe weather events"
author: "James Whitehead"
date: "13 December 2016"
output: html_document
---

##Data Processing

#Loading the data:

```{r}
library(dplyr)
library(plyr)
```

```{r LoadData, cache=TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileURL, "StormData.csv.bz2")
StormData <- read.csv("StormData.csv.bz2")
StormData$BGN_DATE <- as.Date(StormData$BGN_DATE, "%m/%d/%Y")
```

#Processing the data

Only data since Jan 1996 are considered in order to include all event types in the analysis, and avoid skewing towards those recorded in ealier years. Events were also removed where no injuries, fatalities or damage were incurred.
```{r ProcessData, cache=TRUE}
StormData_trimmed <- StormData[StormData$BGN_DATE > as.Date("1996-01-01"), ]
anyCost <- with(StormData_trimmed, FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0)
StormData_trimmed <- StormData_trimmed[anyCost, ]
StormData_trimmed <- subset(StormData_trimmed, select=c("BGN_DATE", "STATE", "EVTYPE", "MAG",
        "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))
```
The exponents of crop and property damage are converted to values, and used to multiply to get total crop and property damage, which are added together to get total damage in USD.
```{r}
PropExp <- as.numeric(as.character(mapvalues(StormData_trimmed$PROPDMGEXP,
        c("K", "M", "B", ""), c("1000", "1000000", "1000000000", "1"))))
CropExp <- as.numeric(as.character(mapvalues(StormData_trimmed$CROPDMGEXP,
        c("K", "M", "B", ""), c("1000", "1000000", "1000000000", "1"))))
StormData_trimmed$TOTALDMGVAL <- with(StormData_trimmed,
        (PROPDMG * PropExp) + (CROPDMG * CropExp))
```
The same event type can be entered differently for different rows; here the names of some of the more common events are renamed so they can be grouped properly.
```{r SortEventType}
StormData_trimmed$EVTYPE <- toupper(StormData_trimmed$EVTYPE)
StormData_trimmed$EVTYPE[grep("^TSTM|^THUNDERSTORM|^ TSTM", StormData_trimmed$EVTYPE)] <-
        "THUNDERSTORM WIND"
StormData_trimmed$EVTYPE[grep("FLASH", StormData_trimmed$EVTYPE)] <- "FLASH FLOOD"
StormData_trimmed$EVTYPE[grep("COASTAL FLOOD|CSTL", StormData_trimmed$EVTYPE)] <-
        "COASTAL FLOOD"
StormData_trimmed$EVTYPE[grep("STRONG WIND", StormData_trimmed$EVTYPE)] <- "STRONG WIND"
StormData_trimmed$EVTYPE[grep("FIRE", StormData_trimmed$EVTYPE)] <- "WILDFIRE"
StormData_trimmed$EVTYPE[grep("RIP CURRENT", StormData_trimmed$EVTYPE)] <- "RIP CURRENT"
StormData_trimmed$EVTYPE[grep("WINTER WEATHER", StormData_trimmed$EVTYPE)] <- "WINTER WEATHER"
StormData_trimmed$EVTYPE[grep("EXTREME COLD|EXTREME WINDCHILL", StormData_trimmed$EVTYPE)] <-
        "EXTREME COLD/EXTREME WINDCHILL"
StormData_trimmed$EVTYPE[grep("HIGH WIND", StormData_trimmed$EVTYPE)] <- "HIGH WIND"
StormData_trimmed$EVTYPE[grep("STRONG WIND", StormData_trimmed$EVTYPE)] <- "STRONG WIND"
StormData_trimmed$EVTYPE[grep("HURRICANE", StormData_trimmed$EVTYPE)] <- "HURRICANE/TYPHOON"
StormData_trimmed$EVTYPE[grep("HEAVY SNOW|EXCESSIVE SNOW", StormData_trimmed$EVTYPE)] <-
        "HEAVY SNOW"
StormData_trimmed$EVTYPE[grep("HEAT WAVE", StormData_trimmed$EVTYPE)] <- "HEAT"
StormData_trimmed$EVTYPE[grep("STORM SURGE", StormData_trimmed$EVTYPE)] <- "STORM SURGE"
StormData_trimmed$EVTYPE[grep("FROST|FREEZE", StormData_trimmed$EVTYPE)] <- "FROST/FREEZE"
StormData_trimmed$EVTYPE[grep("HEAVY RAIN", StormData_trimmed$EVTYPE)] <- "HEAVY RAIN"
```

